---
title: "BIOSTAT 702 Homework 1 - Due 09/25/2022"
author: "Will Powell"
date: "09/19/2022"
output:
  bookdown::word_document2:
    reference_docx: template_README_and_styles_3.docx
    number_sections: no
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, echo=FALSE, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, fig.width=5, fig.height=5)


# ######################################################################### #
##                                                                         ##
##  Author: tina.davenport@duke.edu                                        ##
##  Program: BIOSTAT702_HW02_Lname_Fname.rmd                               ##
##  Purpose: This program drafts the homework and responses                ##
##                                                                         ##
##  Input files:  [list data or source files]                              ##
##                                                                         ##
##  Output files: BIOSTAT702_HW02_Lname_Fname.docx - Word file of answers  ##
##                to homework.                                             ##
##                                                                         ##
##  Change Log:                                                            ##
##  09/18/2022 File started                                                ##
##                                                                         ##
# ######################################################################### #

library(tidyverse)
library(lubridate)
library(knitr)

options(knitr.kable.NA="")
options(max.print = .Machine$integer.max)  # don't limit console printing

rm(list=ls())
```


# Question (6pts)

Let $Y_1,Y_2,\ldots, Y_n$ be an iid sample of $Y \sim N(\mu_Y, \sigma_Y^2)$. Let $X=\bar{Y}=(1/n)\sum_i Y_i$.

<br>

A. (1pt) What is $f_X(\cdot)$, then distribution of $X$ (in terms of $\mu_Y$ and $\sigma_Y^2$).

The distributin of X is $X \sim N(\mu_Y, \sigma_Y^2 / n)$


<br>

B. (2pts) Let $X_1,X_2,\ldots, X_m$ be a random sample of $Y \sim f_X(\cdot)$, where $f_X(\cdot)$ is the distribution you just specified in part (A). Let $W=\bar{X}=(1/m)\sum_i X_i$. What is $f_W(\cdot)$, the distribution of $W$?

$W \sim N(\mu_Y, \sigma_Y^2 / mn)$ 


<br>

C. (2pts) Let $V=aW+b$. What is $f_V(\cdot)$, the distribution of $V$?

$V \sim N(a\mu_Y + b, a^2\sigma_Y^2 / mn)$ 


<br>


# Question (10pts)

Let $Y_1,Y_2,\ldots, Y_{50}$ be an iid sample of $Y \sim N(\mu, \sigma^2)$. Let $\bar{Y}=(1/50)\sum_i Y_i$ be the sample mean, $S^2=\{1/(50-1)\} \sum_i (Y_i-\bar{Y})^2$ be the sample variance, and let $T=(\bar{Y} - \mu) / (S/ \sqrt{50})$.

<br>

A. (2pts) What is $f_T(\cdot)$, the sampling distribution of $T$?

$T$ is from a $t$ distribution of $49$ degrees of freedom.


<br>

B. (4pt) Describe $f_T(-0.65)$ in words. Calculate $f_T(-0.65)$. Is this quantity of practical use? Explain why or why not.

```{r, echo=FALSE}
# recall the built in d-, p-, q-, and r- functions for a variety of 
# distributions
x <- dt(-.65, 49)
```

$f_T(-0.65)$ is the value of the height of the PDF, or density function, at the value $x=-.65$. Its value is not really of practical use since $t$ is a continuous distribution and we should instead look to calculate the probability that its value lies in a given range, not the exact probability. The value of $f_T(-0.65)$ is `r x`.


<br>

C. (4pts) Let $F_T(\cdot)$ be the CDF of $T$. Describe $F_T(-0.65)$ in words. Calculate $F_T(-0.65)$. Is this quantity of practical use? Explain why or why not.

```{r, echo=FALSE}
# recall the built in d-, p-, q-, and r- functions for a variety of 
# distributions
y <- pt(-.65, 49)
```

$F_T(-0.65)$, the CDF of $T$, is the probability that the value of $T$ is less than or equal to $-.065$. Its value is $F_T(-0.65) =$ `r y`.
This is of much more practical use, as it is a tangible quantity that represents the probability that $T$ falls in a range of values and for different values can be used for help in determining confidence intervals. 


<br>


# Question (8pts)

Let $Y \sim f_Y(\cdot)= \textrm{Bin}(15, 0.7)$.

<br>

A. (4pt) Describe $f_Y(11)$ in words. Calculate $f_Y(11)$. Is this quantity of practical use? Explain why or why not.

```{r, echo=FALSE}
# recall the built in d-, p-, q-, and r- functions for a variety of 
# distributions
 z <- dbinom(11, 15, 0.7)


```

$f_Y(11)$ is the probability of observing $11$ successes in $15$ trials where the probability of success is $0.7$. $f_Y(11) =$ `r z`. Since this is a discrete distribution, it is of practical use as this probability corresponds to the tangible quantity of observing this many successes in $15$ trials.


<br>

B. (4pts) Let $F_Y(\cdot)$ be the CDF of $Y$. Describe $F_Y(11)$ in words. Calculate $F_Y(11)$. Is this quantity of practical use? Explain why or why not.

```{r, echo=FALSE}
# recall the built in d-, p-, q-, and r- functions for a variety of 
# distributions
a <- pbinom(11, 15, 0.7)

```

$F_Y(11)$ is the probability of observing $11$ or less successes in $15$ trials. $F_Y(11) =$ `r a`. This is also of practical use as we can use it to determine the probability of observing more or less success than a set number in a set number of trials.


<br>


# Question (10pts)

Suppose the number of apneic episodes per hour for individuals being treated with an experimental drug for sleep apnea is $N(1,0.7)$.

A. (2pts) What is the probability that a person chosen at random from this population will have more than 3 episodes per hour?

```{r, echo=FALSE}
# recall the built in d-, p-, q-, and r- functions for a variety of 
# distributions
?dnorm
a <- 1-pnorm(3, 1, sqrt(0.7))
```

P(more than 3 episodes per hour) $=$ `r a`


<br>

B. (2pt) What is the probability that a person chosen at random will have between 1 and 3 (inclusive) episodes per hour?

```{r, echo=FALSE}
# recall the built in d-, p-, q-, and r- functions for a variety of 
# distributions
b <- pnorm(3, 1, sqrt(0.7)) - pnorm(1, 1, sqrt(0.7))
```

P(between 1 and 3 episodes per hour) $=$ `r b`


<br>

C. (2pt) What is the 11th percentile of apneic episodes per hour?

```{r, echo=FALSE}
# recall the built in d-, p-, q-, and r- functions for a variety of 
# distributions
c <- qnorm(.11, 1, sqrt(0.7))
```

11 percentile is `r c` apneic episodes per hour


<br>

D. (4pts) Based on your answer in (C), what is an issue with assuming that apneic episodes per hour is normally distributed? 

The issue is that this value is negative, and it is impossible to have a negative number of apneic episodes per hour. This any values in these lower percentiles are unrealistic. Another distribution that does not include negative better values may be a better approximation.


<br>


# Question (16pts)

Review the description of the SWISS data from previous homework. Create the *cleaned* SWISS data using the code below.

```{r}
SWISS <- read.csv(file="SWISSlong.csv", na.strings="")

SWISSwide <- SWISS %>%
  filter(Year %in% c(68, 78)) %>%  # select the years of interest
  select(partID, group, Year, eGFR)  %>%  # choose a few columns
  reshape(v.names="eGFR",    # variable that will be multiple columns
      timevar="Year",        # variable indicating replications
      idvar="partID",        # variable indicating unique participants
      direction="wide") %>%  # which direction to be converted to
  mutate(eGFRchg=eGFR.78-eGFR.68)  # calculate change in eGFR
```

<br>

A. (6pts) Is the change in eGFR from year 1968 to year 1978 approximately normally distributed? Justify your answer with numerical summaries and plots (including a Q-Q plot).

```{r, echo=FALSE}
# see ?qqnorm and ?qqline for help with a Normal Q-Q plot 
library(ggplot2)

hist(SWISSwide$eGFRchg, main = 'Histogram of Change in eGFR from 1968 to 1978', col='Blue')

qqnorm(SWISSwide$eGFRchg, main= 'Normal QQ Plot of change in eGFR from 1968 to 1978')
qqline(SWISSwide$eGFRchg, col = "steelblue", lwd = 2)

ggplot(data=SWISSwide, aes(x=SWISSwide$eGFRchg), na.rm=TRUE) + 
  geom_density(alpha=0.6) 

```

Change in eGFR has a mean of `r mean(na.omit(SWISSwide$eGFRchg))` and a standard deviation of `r sd(na.omit(SWISSwide$eGFRchg))`. When looking at the above plots, we see that the data appear to be approximately normally distributed. 


<br>

B. (6pts) Regardless of you answer in (A) construct 95% CIs BY HAND (i.e., don't use ?t.test) for the population mean eGFR at year 1968, year 1978, and the change in eGFR.

```{r, echo=FALSE}
# you may check your answer against ?t.test, but I want to see the formula
# (or you may write your own function!) coded out here.


CI <- function(data){
  data <- na.omit(data)
  n <- length(data)
  y_bar <- mean(data)
  lower <- y_bar - qt(.975, n-1) *sd(data)/sqrt(n)
  upper <- y_bar + qt(.975, n-1)*sd(data)/sqrt(n)
  return(c(lower, upper))
}

CI_68 <- CI(SWISSwide$eGFR.68)
CI_78 <- CI(SWISSwide$eGFR.78)
CI_chg <- CI(SWISSwide$eGFRchg)

df <- data.frame(CI_68, CI_78, CI_chg)
row.names(df) <- c('lower', 'upper')
df <- t(df)
row.names(df) <- c('1968 CI for mean eGFR', '1978 CI for mean eGFR', 'CI for change in mean eGFR')
df <- data.frame(df)
df
```

The confidence intervals are printed above

<br>


C. (4pts) Interpret the CI for change in eGFR in a way that a clinician (non-statistician) would understand.


We are 95% confidente that the true value for the change in mean eGFR is between -10.14 mL/min and -4.83 mL/min. This doesn't include zero, so we are confident there is a significant change in this time period

<br>




# Session Info

```{r}
sessionInfo()
```

<br>

--- END OF REPORT ---